Relational Parametricity for Algebraically Indexed Types

### 2012-05-29 paper/section-general-framework.tex
Fix some texing errors in the proof of the abstraction theorem

The following lemma states that the relation interpretation of types
that we have defined in this section behaves well: the first part of
the lemma states that two types that are judgmentally equal are given
equal relational interpretations, and the second part states that
substitution of index terms in types can be interpreted by the
composition of relational environments with simultaneous
substitutions. Note that both parts of the lemma depend on
\lemref{lem:tyeqsubst-erasure} to be ``well-typed''.
\begin{lemma}\label{lem:tyeqsubst-relational}
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$, then for all $\rho \in
    \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}{\rho} =
    \rsem{B}{\relEnv{E}}{\rho}$;
  \item If $\Delta \vdash A \isType$ and $\sigma : \Delta' \Rightarrow
    \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
    $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}(\rho
    \circ \sigma)$.
  \end{enumerate}
\end{lemma}

@@

\begin{proof}
  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
  three interesting cases follow:
  \begin{description}
  \item[Case \TirName{TyEq}] This case follows directly from
    \lemref{lem:tyeqsubst-relational}.
  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
    $A = \forall i\mathord:s. A'$. We know from the induction
    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}\rho'$,
    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}\rho'$. We are given $\rho \in
    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
    \rsem{\pi_{i:s}^*\Gamma}\rho''$ by \lemref{lem:tysubst-rel}. So we
    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}\rho''$, and thus
    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
      i.M'}\eta_2) \in \rsem{\forall
      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
    know that for all $\rho' \in \extends{\rho}{i:s}$,
    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}\rho'$. If we let $\rho' = \rho \circ
    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
    by \lemref{lem:tysubst-rel}, conclude that $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
    as required.
  \end{description}
  The remaining cases are standard for proofs of the fundamental lemma
  of logical relations for simply-typed systems.
\end{proof}

>>>>

The following lemma states that the relation interpretation of types
that we have defined in this section behaves well: the first part of
the lemma states that two types that are judgmentally equal are given
equal relational interpretations, and the second part states that
substitution of index terms in types can be interpreted by the
composition of relational environments with simultaneous
substitutions. Note that both parts of the lemma depend on
\lemref{lem:tyeqsubst-erasure} to be ``well-typed''.
\begin{lemma}\label{lem:tyeqsubst-relational}
  \begin{enumerate}
  \item If $\Delta \vdash A \equiv B \isType$, then for all $\rho \in
    \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}{\rho} =
    \rsem{B}{\relEnv{E}}{\rho}$;
  \item If $\Delta' \vdash A \isType$ and $\sigma : \Delta \Rightarrow
    \Delta'$, then for all $\rho \in \relEnv{E}(\Delta)$,
    $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}(\rho
    \circ \sigma)$.
  \end{enumerate}
\end{lemma}

@@

\begin{proof}
  By induction on the derivation of $\Delta; \Gamma \vdash M : A$. The
  three interesting cases follow:
  \begin{description}
  \item[Case \TirName{TyEq}] This case follows directly from
    \lemref{lem:tyeqsubst-relational}.
  \item[Case \TirName{UnivAbs}] In this case, $M = \Lambda i.M'$ and
    $A = \forall i\mathord:s. A'$. We know from the induction
    hypothesis that for all $\rho' \in \relEnv{E}(\Delta, i:s)$ and
    $(\eta_1,\eta_2) \in \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho'}$,
    we have $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{\rho'}$. We are given $\rho \in
    \relEnv{E}(\Delta)$, $(\eta_1,\eta_2) \in
    \rsem{\Gamma}{\relEnv{E}}\rho$ and $\rho'' \in
    \extends{\rho}{i:s}$. Since $\rho'' \circ \pi^*_{i:s} = \rho$, we
    have $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}\rho$ implies
    $(\eta_1,\eta_2) \in \rsem{\Gamma}{\relEnv{E}}{(\rho'' \circ
      \pi^*_{i:s})}$, and therefore $(\eta_1,\eta_2) \in
    \rsem{\pi_{i:s}^*\Gamma}{\relEnv{E}}{\rho''}$ by \lemref{lem:tysubst-rel}. So we
    can apply the induction hypothesis to get $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{A'}{\relEnv{E}}{\rho''}$, and thus
    $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda i.M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}\rho''$, by the definition of $\tmSem{\Lambda
      i. M'}$. Hence $(\tmSem{\Lambda i.M'}\eta_1,\tmSem{\Lambda
      i.M'}\eta_2) \in \rsem{\forall
      i\mathord:s. A'}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{UnivApp}] In this case, $M = M'[e]$ and $A =
    (\id_\Delta, e)^*A'$. By the induction hypothesis, and the
    definition of $\rsem{\forall i\mathord:s.A'}{\relEnv{E}}\rho$, we
    know that for all $\rho' \in \extends{\rho}{i:s}$,
    $(\tmSem{M'}\eta_1, \tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{\rho'}$. If we let $\rho' = \rho \circ
    (\id_\Delta, e)^*$, we know that $\rho' \in \extends{\rho}{i:s}$,
    because $\rho' \circ \pi^*_{i:s} = \rho \circ (\id_\Delta, e)^*
    \circ \pi^*_{i:s} = \rho$. Hence we may deduce that
    $(\tmSem{M'}\eta_1,\tmSem{M'}\eta_2) \in
    \rsem{A'}{\relEnv{E}}{(\rho \circ (\id_\Delta,e)^*)}$ and thence,
    by \lemref{lem:tysubst-rel}, conclude that $(\tmSem{M'}\eta_1,
    \tmSem{M'}\eta_2) \in \rsem{(\id_\Delta,e)^*A'}{\relEnv{E}}\rho$,
    as required.
  \end{description}
  The remaining cases are standard for proofs of the fundamental lemma
  of logical relations for simply-typed systems.
\end{proof}
